---
title: "lab 5: Chick Weight dataset visualisation"
output: html_document
---

The aim of this lab is to understand **split-apply-combine pattern** and apply the pattern to `ChickWeight dataset`.

```{r}
#install.packages("datasets") 
library(datasets)
#data()
```

Now look at ChickWeight dataset. This gives the data from the experiment described in the lecture on "Data Wrangling".

The `ChickWeight data` frame has 578 rows and 4 columns. It has 4 variables:

- weight - a numeric value giving the body weight of the chick (gm). 
- Time - the number of days since birth when the measurement was made. 
- Chick - a unique identifier for the chick. 
- Diet - indicating which experimental diet the chick received.

```{r}
head(ChickWeight)
str(ChickWeight)
names(ChickWeight)
```

It’s annoying that some of the names are capitalized and some are not. We can fix this and have everything lower case: 

```{r}
chickweight <- ChickWeight
names(chickweight) <- tolower(names(chickweight))
#?tolower
str(chickweight)
names(chickweight)
```

Good! 
A graphical overview of the dataset can be done:

```{r}
boxplot(weight~time, data = chickweight, 
        xlab="time", ylab="weight")
#?boxplot
```

## Subset data

If we only want the weights of each chick on day 21, we can subset the data:

```{r}
cw21 <- subset(chickweight, time == 21)
summary(cw21)

#?summary
#?subset
#is.na(chickweight)
#?is.na
#anyNA(chickweight, recursive = FALSE)
```

To make `cw21` we used conditional subsetting (time == 21). Available comparison operators are:

- < less than
- ( > greater than)
- == equal to
- <= less than or equal to
- (>= greater than or equal to)
- != NOT equal to (! symbol indicates negation)
- is.na(x) tests if x has missing values

Logical operators to combine expressions are also available:

- & logical AND
- ( | logical OR)
- ! logical NOT (negation)

**Sometimes, we want to subset based on multiple variables,** 

for example, we canfind how many chicks on diet 1 have weights over 250 on day 21:

```{r}
cw21.subset2 <- cw21[(cw21$diet == "1" & cw21$weight >= 250), ]
cw21.subset2
```

## Using plyr package

Now let’s use `plyr`:

```{r}
#?plyr
#install.packages("plyr")
library(plyr) 
library(ggplot2)
```

#### Let’s plot the weight of each chick as a function of time. 

This requires a little thought. 

- x will be time; 
- y will be weight. 

But we want a separate line for each chick - but it might be best if the lines are coloured according to diet. 

There are lots of chicks so we don’t want a different colour for each chick - there will be too many, and the whole point is to distinguish the effects of the diets. 

```{r}
p <- ggplot(chickweight,aes(x=time,y=weight, 
                            colour=diet, 
                            group=chick)) + geom_line()
p
#?ggplot
```

What do you see? 

Let’s summarise to mean weight for each diet on each day.

```{r}
chicksummary1 <- ddply( chickweight, .(diet,time), summarise,
                        meanweight=mean(weight) )
str(chicksummary1) 
names( chicksummary1)
chicksummary1
#?ddply
```

Now let’s plot it:

```{r}
p <- ggplot( chicksummary1,aes(x=time,y=meanweight,colour=diet)) +geom_line()
p
```

What do you see?

**Try using a log scale for the y axis** 

- what do you see? 
- Is the chick growth exponential for part of the time? 
- Is this reasonable?

```{r}
p + scale_y_log10()+ylab("log of meanweight")
```

## Challenges:

#### 1. Use `ddply` to produce other summaries.

- How many chicks were fed each diet?
- How many chicks, fed each diet, died early? Which day did they die?

**How many chicks were fed each diet?**

```{r}
chicksummary2a <- ddply(chickweight, .(diet), summarise,
                         numberofchick=max(chick))
chicksummary2a
```

**How many chicks, fed each diet, died early? Which day did they die?**

```{r}
chicksummary1b <- ddply( chickweight, .(diet), summarise,
             lastday=max(time),numberofchick=max(chick))
chicksummary1b
```

#### 2. Create an individual facet for each diet within the dataset to show the relationship between times and weights.

```{r}
p <- ggplot(chickweight, aes(x=time, y=weight, colour=diet)) + geom_point()

p+facet_grid(diet~.)

# ?geom_point()
# ?facet_grid
```

#### 3. Do the line plots we have done show that one diet is better than another? 

(Hint: they don’t ! Not quite.) What would be a better plot to compare how heavy chicks raised on different diets became? 

(Hint: boxplots. Look up notched boxplots on page 133 in the “R Graphics Cookbook” by Winston Chang. )

```{r}
ggplot(chickweight, aes(x=time,y=weight, 
                        colour=diet, group=chick))+ geom_boxplot()+ scale_x_continuous(breaks = NULL)

# ?geom_boxplot()
# ?scale_x_continuous
```

````

````
