---
title: "lab 0"
output: html_document
---

## Lab 0: Introduction to R

- Type 'demo()' for some demos, 'help()' for on-line help, or
- 'help.start()' for an HTML browser interface to help.
- Type 'q()' to quit R.

## Basic Commands

Data frame

```{r}
x <- 17.3
x
7*x
```

```{r}
remove(x) # x disappears
```
`getwd()` - Prints your working directory. 

setwd( “the directory you want here” )

Vectors (one dimensional arrays) are made like this: 

```{r}
v <- c( 1.1, 2.2, 17.3, -23, 4, 0)
v
v[1]
v[2:5]
```

Logical (Boolean) vectors are useful. 

Note that this is a Boolean vector, and TRUE and FALSE are printed (and entered) in capitals.

Logical vectors are very useful for selecting subsets of elements of a vector. 


```{r}
v
boolv <- v < 3
boolv 
# This returns a vector of all elements of v that are less than 3. 
v[boolv]
```

We can write an expression that gives a logical vector inside the subscript brackets: 

```{r}
v
v[ v < 3 ]
```

## Lists:

Lists can contain elements of different types.

Giving names to elements of their lists. The advantage of doing this is that you can refer to elements of a list by name, without having to remember what position they are:

```{r}
mylist <- list( alf=34, bert=c(2,3,4,5), 17.3, pain="wisdom tooth")
mylist
```

```{r}
mylist$alf
mylist$pain
```

Alternatively, you can refer to them by position: 

```{r}
mylist[2]

class( mylist )
class( mylist[2] )
```

We see that mylist[2] is actually a one element list. To extract an individual element from a list by its index, we use

```{r}
mylist[[2]]

class( mylist[[2]] )
class( mylist $bert )
```

We can remove an element from a list by: 

```{r}
mylist$bert <- NULL
mylist
mylist$bert
```

## Data Frames

A data frame is a class, designed to hold data in a convenient way. It is implemented as a list of vectors, which should be of equal lengths. 

Classes in R are implemented as lists.

```{r}
### Data Frames
mydf <- data.frame(row.names= c("alf","bert", "r2d2"),
                   sex=c("M","F","R"),
                   age=c(20,21,3) )
mydf
```
We can use a logical vector (which we can compute in any way we like) to select a subset of rows.

```{r}
mydf[c(FALSE,TRUE,TRUE), ]
```
This selects the vector of ages.

```{r}
mydf$age
```

```{r}
class( mydf)
class(mydf$sex)
```

A character object is used to represent string values in R. We can convert character into factor:

Each string (in this case “M”, “F”, 
or “R”) is encoded as a number (the numbers 1, 2, and 3), and there is a separate (invisible) array that stores the strings.

```{r}
class(as.factor(mydf$sex))
```

## Functions

Write your own function “area” to calculate the area of a circle with a given radius:

```{r}
# Functions
area <- function(radius) {
  pi * radius^2
}
area(10)
```

### The Pipe: a tidy operator for functions.

%>% is a popular new operator, increasingly used to clean big, messy data sets.

```{r}
# The Pipe: a tidy operator for functions
library(tidyr) # %>% originally from library(magrittr)
x <- rnorm(100, mean = 100) # # 100 random normal draws
x
x %>% range
```

It is often used when several functions need to be applied in a row (where scale is a function to standardize a data).

```{r}
x %>% scale %>% abs %>% range %>% round(1)
```

```{r}
y <- x %>% scale %>% abs %>% range %>% round(1)
y
```

## Loading data

```{r}
Auto <- read.table("Auto.data")
dim(Auto)
#fix(Auto)
head(Auto, 10)
```


```{r}
#Auto<-read.table("Auto.data", sep=",",header=T, na.strings="?", quote="")
Auto <- read.table("Auto.data", header=T, na.strings="?")
#fix(Auto)
head(Auto, 10)
```


```{r}
Auto<-na.omit(Auto)
dim(Auto)
```
Once the data are loaded correctly, we can use names() to check the variable names.

```{r}
names(Auto)
```

## Writing scripts in R

```{r}
### Writing scripts in R
if (2 > 0) {
  i <- 2
  print (i)
}
```
If the name of this file is `script.R`, you can run all these commands in one go as: source(“script.R”) Files used in this way are called scripts.

## Saving plots in R

It is possible to save the graph using R codes as followings:

```{r}
## Saving plots in R
jpeg("plot1.jpg")
cars<-c(1, 3, 6, 4, 9)
plot(cars, type="o", col="blue")
title(main="Autos", col.main="red", font.main=4)
dev.off()

```



````

````







