---
title: "lab 1 - Exercises"
output: html_document
---

### For this exercise we are using `the iris data set` available in R.

#### 1. Make a scatter plot of sepal length vs. petal length for the three species. Make a single plot that shows the data for all three species at once, in different colours.

```{r}
library(ggplot2)
names(iris)
head(iris, 10)

ggplot(iris, aes(x=Sepal.Length, y=Petal.Length, color=Species)) + geom_point()

ggplot(iris, aes(x=Sepal.Width, y=Petal.Width, color=Species)) + geom_point()
```

#### 2. Now try to show the data for all three species at once, in different colours and shapes.

```{r}
ggplot(iris, aes(x=Sepal.Length, y=Petal.Length, shape=Species, color=Species)) + 
  geom_point()
```

#### 3. Finally, add a fitted least squares line (straight black line) to each group of points.

Hint: use geom_smooth, which adds a smoother (possibly a least squares line, possibly a 
smooth curve fit to the data, etc.). The argument method = lm specifies a line fitted by least 
squares, and the argument se = FALSE suppresses the default display of a confidence band 
around the line or curve which was fit to the data.

```{r}
ggplot(iris, aes(x=Sepal.Length, y=Petal.Length, color=Species)) + geom_point() + 
  geom_smooth(aes(group=Species), method='lm', color='black', formula=y~x, se=F)
```

----

----

